SFJ: An implementation of Semantic Featherweight Java

### 2020-02-25 conference/paper.tex
reduce spacing in algorithms, add further explanations and details

@@

However, these benefits were only presented and exemplified at a theoretical level and not exploited practically. 
At last, in this work we have implemented SFJ---Semantic Featherweight Java, which features structural subtyping and boolean types as well as nominal subtyping, typical of FJ itself.
The benefits of SFJ, also illustrated in the accompanied video, show how static typechecking of boolean types and semantic subtyping gives higher guarantees of program correctness, more flexibility and compactness of program writing.
\keywords{Nominal subtyping \and Structural subtyping  \and Semantic Featherweight Java \and Object-oriented languages \and Boolean types \and Type theory.}
\end{abstract}

\section{Introduction}
There are two approaches to defining subtyping relations: the \emph{syntactic} and the \emph{semantic} approach.
In the latter, one starts by defining a model of the programming language at hand and an interpretation of its types as subsets of such model.
The syntactic subtyping is more mainstream in programming languages design and implementation, and the relation is given by means of a set of formal deductive subtyping rules.
The semantic subtyping is less know and the relation is given by inclusion of sets denoting types. One starts from a formal model of the programming language at hand and defines an interpretation, or semantics, of types as subsets of this model; subtyping naturally is defined as inclusion of these sets, hence the semantic adjective attached to it.
Orthogonally, especially for object-oriented languages, there are two approaches to defining subtyping relation: the \emph{nominal} and the \emph{structural} approach.
The nominal subtyping relation is based on \emph{declarations} by the developer and is \emph{name}-based, hance the adjective nominal: $A$ is a subtype of $B$ if and only if it is declared to be so, that is if the class $A$ extends (or implements) the class (or interface) $B$.
The structural subtyping relation instead is based on the \emph{structure} of a class, its fields and methods: a class $A$ is a subtype of a class $B$ if and only if the fields and methods of $A$ are a superset of the fields and methods of $B$, and their types in $A$ are subtypes of the types in $B$.
At this point it is not surprising that the nominal subtyping aligns well with the syntactic approach and hence is more well-known than the structural subtyping which aligns well with the semantic approach.

@@

The approach proposed by Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} allows for the class hierarchy to be defined in such a natural way without needing to change it, but on the other hand it does not require handling exceptions at run time,  rather just static typechecking.
To achieve this, the framework combines boolean types with semantic subtyping: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} and not \emph{Triangle} thus allowing the typechecker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\syndecl{Polygon}{Object} {\cdots}
\\
\syndecl{Triangle}{Polygon} {\cdots}
\\
\syndecl{Rectangle}{Polygon} {\cdots}
\\
\syndecl{Rumble}{Polygon} {\cdots}
\\
\hspace{2.5cm} \ldots
\\
\syndecl{Diagonal}{Object} {
	\\
	\qquad \mathbf{real}\ diagonal(\textit{Polygon} \textbf{ and not }  \textit{Triangle} \ p)\{\ldots\}
	\\
}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}
The combination of boolean types and semantic subtyping gives a more elegant solution, as illustrate by the \emph{Polygons}---of course a toy example used only as a proof of concept of the approach---however, this approach is not yet practical as there are no boolean types or semantic subtyping in Java, hence its benefits unfortunately remained theoretical.

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
Since we want our types to represent sets of values, we restrict our types to finite trees whose 
leaves are constants with no cycles. For example a recursive type $\alpha = [a : \alpha]$ would
be an infinite tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)). Similarly the types $\alpha = 
[b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.

Therefore, to not allow this, when processing the AST of a program, we do not allow the type of 
the class we are defining to be a field and we mark any classes with only basic types in it's fields
as \textit{resolved} and otherwise if it contains a class type, we mark it as \textit{unresolved}. 
After processing the whole AST, we perform the following algorithm to decide whether the type 
definitions in the program are valid.

\begin{verbatim}
boolean resolutionOccured = false
do 
    for class that is unresolved:
        boolean resolved = true

        for field that contains a class type:
            if class type is not resolved 
                resolved = false

        if resolved is true
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

if all classes are not resolved
    program contains invalid type definition
\end{verbatim}

Given that we now know that all the types in the program are valid, we can create the subtyping 
relation which will be a map of types to a set of its subtypes. For every program, we assume the 
initial subtyping relation for all of our basic types. We would like to highlight that Int is not
a subtype of Float as it cannot represent the whole domain of Int accurately therefore the domain is 
not fully contained in Float. Similarly so for Double and Long.

\begin{flalign*}
    Boolean &= \left\{Boolean \right\} &\\
    Double  &= \left\{Double, Float, Int, Short, Byte\right\} &\\
    Float &= \left\{Float, Short, Byte\right\} &\\
    Long &= \left\{Long, Int, Short, Byte\right\} &\\
    Int &= \left\{Int, Short, Byte\right\} &\\
    Short &= \left\{Short, Byte\right\} &\\
    Byte &= \left\{Byte\right\} & 
\end{flalign*}

Knowing that all our types are finite trees with leaves as constants and given this initial
relation, we can now confidently create a subtyping relation for all class types using the following
algorithm.

\begin{verbatim}
function generateRelation(classes):
    List<class> untyped = []

    for class in classes:
        if addClass(class) is false:
            untyped.add(class)

    if untyped is not []:
        generateRelation(untyped)

function addClass(class):
    for existing class type in relation:
        if tryAddSubtype(class, existingClass) is false:
            return false

        tryAddSubtype(existingClass, class)

        add class to its own subtype relation

function tryAddSubtype(class, other):
    boolean flag = true

    for field in class:
        if field contains type not in relation:
            return false

        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false

    for method in class:
        if method contains type not in relation:
            return false

        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false

    if flag == true:
        add class to other subtype relation
\end{verbatim}

The following bibliography provides
a sample reference list with entries for journal
articles~\cite{Dardha2017}.

>>>>

\makeatletter
\def\verbatim@font{\linespread{0.75}\normalfont\ttfamily}
\makeatother

@@

However, these benefits were only presented and exemplified at a theoretical level and not exploited practically.
At last, in this work we have implemented SFJ---Semantic Featherweight Java, which features structural subtyping and boolean types as well as nominal subtyping, typical of FJ itself.
The benefits of SFJ, also illustrated in the accompanied video, show how static type checking of boolean types and semantic subtyping gives higher guarantees of program correctness, more flexibility and compactness of program writing.
\keywords{Nominal subtyping \and Structural subtyping  \and Semantic Featherweight Java \and Object-oriented languages \and Boolean types \and Type theory.}
\end{abstract}

\section{Introduction}
There are two approaches to defining subtyping relations: the \emph{syntactic} and the \emph{semantic} approach.
In the latter, one starts by defining a model of the programming language at hand and an interpretation of its types as subsets of such model.
The syntactic subtyping is more mainstream in programming languages design and implementation, and the relation is given by means of a set of formal deductive subtyping rules.
The semantic subtyping is less know and the relation is given by inclusion of sets denoting types. One starts from a formal model of the programming language at hand and defines an interpretation, or semantics, of types as subsets of this model; subtyping naturally is defined as inclusion of these sets, hence the semantic adjective attached to it.
Orthogonally, especially for object-oriented languages, there are two approaches to defining subtyping relation: the \emph{nominal} and the \emph{structural} approach.
The nominal subtyping relation is based on \emph{declarations} by the developer and is \emph{name}-based, hence the adjective nominal: $A$ is a subtype of $B$ if and only if it is declared to be so, that is if the class $A$ extends (or implements) the class (or interface) $B$.
The structural subtyping relation instead is based on the \emph{structure} of a class, its fields and methods: a class $A$ is a subtype of a class $B$ if and only if the fields and methods of $A$ are a superset of the fields and methods of $B$, and their types in $A$ are subtypes of the types in $B$.
At this point it is not surprising that the nominal subtyping aligns well with the syntactic approach and hence is more well-known than the structural subtyping which aligns well with the semantic approach.

@@

The approach proposed by Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} allows for the class hierarchy to be defined in such a natural way without needing to change it, but on the other hand it does not require handling exceptions at run time,  rather just static type checking.
To achieve this, the framework combines boolean types with semantic subtyping: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} and not \emph{Triangle} thus allowing the type checker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\syndecl{Polygon}{Object} {\cdots}
\\
\syndecl{Triangle}{Polygon} {\cdots}
\\
\syndecl{Rectangle}{Polygon} {\cdots}
\\
\syndecl{Rumble}{Polygon} {\cdots}
\\
\hspace{2.5cm} \ldots
\\
\syndecl{Diagonal}{Object} {
	\\
	\qquad \mathbf{real}\ diagonal(\textit{Polygon} \textbf{ and not }  \textit{Triangle} \ p)\{\ldots\}
	\\
}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}
The combination of boolean types and semantic subtyping gives a more elegant solution, as illustrate by the \emph{Polygons}---of course a toy example used only as a proof of concept of the approach---however, this approach is not yet practical as there are no boolean types or semantic subtyping in Java, hence its benefits unfortunately remained theoretical.

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
The key notion in semantic subtyping is denoting types as sets of values. Since we want these types
to be able to be finitely representable, we restrict our types to finite trees whose
leaves are constants with no cycles. For example a recursive type $\alpha = [a : \alpha]$ would
be an infinite tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)). Similarly the types $\alpha =
[b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.

Therefore, when processing the abstract syntax tree (AST) of a program, we do not allow the type of
the class we are defining to be the type of its own field. Secondly we mark any classes with only
basic types \textit{(bool, int, float ...)} for the types of its fields as \textit{resolved} and otherwise, if it contains a class type, we mark it as \textit{unresolved}. After processing the whole AST, we perform the following algorithm to decide whether the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

if all classes are not resolved:
    program contains invalid type definition
\end{verbatim}

If the types in the program are finite trees finite trees whose leaves are constants with no cycles,
then at each iteration of the algorithm, either all the types are going to be resolved, or we are
going to be able to resolve at least one type. If we do not resolve at least one type and not all 
types are resolved, we know we have encountered a cycle in the type definition.

Given that we now know that all the types in the program are valid, we can create the subtyping
relation which will be a map of types to a set of its subtypes. For every program, we assume an
initial subtyping relation for all of our basic types. We would like to highlight that \textit{Int}
is not a subtype of \textit{Float} as it cannot represent the whole domain of \textit{Int} accurately
and therefore the domain is not fully contained, however this is not the case for \textit{Double} 
and \textit{Int}. Similarly so \textit{Long} is not a subtype of \textit{Double}.
\begin{flalign*}
    Boolean &= \left\{Boolean \right\} &\\
    Double  &= \left\{Double, Float, Int, Short, Byte\right\} &\\
    Float &= \left\{Float, Short, Byte\right\} &\\
    Long &= \left\{Long, Int, Short, Byte\right\} &\\
    Int &= \left\{Int, Short, Byte\right\} &\\
    Short &= \left\{Short, Byte\right\} &\\
    Byte &= \left\{Byte\right\} &
\end{flalign*}

Knowing that all our types are finite trees with leaves as constants and given this initial
relation, we can create a subtyping relation for all class types using the following
algorithm.

\begin{verbatim}
function generateRelation(classes):
    List<class> untyped = []
    for class in classes:
        if addClass(class) is false:
            untyped.add(class)
    if untyped is not []:
        generateRelation(untyped)

function addClass(class):
    for existing class type in relation:
        if checkSuperSet(class, existingClass) is false:
            return false
        checkSuperSet(existingClass, class)
        add class to its own subtype relation

function checkSuperSet(class, other):
    boolean flag = true
    for field in class:
        if field contains type not in relation:
            return false
        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false
    for method in class:
        if method contains type not in relation:
            return false
        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false
    if flag == true:
        add class to other subtype relation
\end{verbatim}

Continuing with the Polygons example, the algorithm would generate the following relation, including all base relations we defined before.
\begin{flalign*}
    Polygon &= \left\{Polygon, Triangle, Rectangle, Rumble \right\} &&\\
    Triangle  &= \left\{Triangle\right\}  
    Rectangle = \left\{Rectangle\right\} 
    Diagonal = \left\{Diagonal\right\} 
\end{flalign*}

Given method \textit{diagonal} from the example, we can see that the result of the set operation on its parameter type would be 
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Rectangle, Rumble \right\} 
\end{flalign*}
Therefore if we try to pass an argument of type $Triangle$ to this method, the type
if not fully contained within the computed set, so it does not type check.

The above subtyping algorithm will find all nominal subtypes, as nominal subtypes inherit all fields
and methods of their super class, so are always guaranteed to be a superset. Similarly, we can check
every pair of types if one is a superset of the other, i.e.\ they share common fields and
methods, and if so, we know it is a structural subtype of the other. For example, the Empty type,
$empty = []$, will have all other classes as structural subtypes. The above algorithm will 
without differentiating, find all nominal and structural subtypes as we always consider all other
classes in the program to be potential subtype candidates.

The argument against structural subtyping comes up here, as a type $coord = [x:int, y:int, z:int]$ will be structurally similar to a type $colour = [x:int, y:int, z:int]$. While they can be used 
interchangeable in our type system, the meaning of the types are completely different so it means
we ignore possible wrong behaviour in our program. We however argue that it is no more unsafe than
nominal subtyping in Java where an overridden method can be defined to perform the opposite
logic to what the super class expects. Therefore both methods leave an expectation on the programmer
to check what they are doing is correct.

Furthermore, we have extended FJ with method types, demonstrated below to show another possible
implementation of the Polygons example, where we only implement the \textit{diagonal} method on 
polygons that support it, and not in the Triangle class. 
$$
\syndecl{Diagonal}{Object} {
	\mathbf{int}\ diagonal(\textit{(getDiagonal: void$\rightarrow$int)}  \ p)\{\ldots\}
}
$$

While we do not generate a relation for each method in all classes as this is computationally 
unnecessary, we can on encountering a method type generate the set of classes which implement this
method and combine their subtype relations using the \textit{or} operator. We can then use the same set inclusion to see if the argument passed to this function type checks.

In the implementation of SFJ, we are liberal in the use of fields and method which might not be 
implemented for all types in the subtyping relation rather than restricting only it to those which
we know are shared by all types. Improper use of this allows for run-time exceptions however it
more closely follows the idea of giving the programmer tools to create more expressive programs
at the cost of checking program correctness such as with the use of structural subtypes.
